Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-02-06 analysis/paper/paper.Rmd
added an abstract and minor editing

---
title: "Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Li-Ying Wang Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could result in far-reaching economic or cultural impacts on local societies. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022]. Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.

@@

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, and culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021]. This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020]. We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture through our case study.

@@

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The common ceramic styles in the early Neolithic period include grayish high-tempered fine ware and reddish brown coarse ware with surface decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Among the ceramics, incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a]. About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016]. The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022]. A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 BP [@Hao2022]. Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

<!--info about Gaoshan-->
To have a comparison, we also look at a sample from Gaoshan, which is located on the southwestern margin of the Chengdu Plain (Figure \@ref(fig:SW-china-map)). The Chengdu Plain is an alluvial plain between the Sichuan Basin and the Tibetan Plateau. Excavations at Gaoshan revealed abundant materials, including pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain, or even earlier than that [@Guedes2013; @Gaoshan2017]. A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows the predominance of domesticated pigs, with a small number of domesticated dogs [@He2020]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation.

Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP, what their proportions are in local culinary practices during the early time are unclear. For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in neighboring sites, such as Gaoshan and Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013]. The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.

@@

knitr::kable(Ch_sherds,
             caption = "Quantitative data of fatty acid (FA) composition (ug) of potsherds from Guijiabao and Gaushan that are named starting with 'SYG' and 'CDG' respectively")
```

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For the sample processing, we cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS. Internal standards were added before and after the total lipid extraction.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). Helium carrier is in constant flow rate of 1.3 ml/min. The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 min. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode. The carrier gas was helium at 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was held at 90°C for 1.3 minutes, and then raised to 320°C at a rate of 5°C/min. The final temperature was held for 22 minutes. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface. The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set from 90 °C for 1.5 minutes, and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.

@@

knitr::kable(quant_FA_tidy,
             caption = "Quantitative data of fatty acid (FA) composition (ug) of potsherds from Guijiabao and Gaushan that are named starting with 'SYG' and 'CDG' respectively")
```

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of ceramic or a different use. Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to the early Neolithic period contains a rich composition of long-chain fatty acids, reflecting good preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes, and a lack of local modern references for animal fats and plants. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

@@

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). However, we are unable to determine whether there is a shift in diet over time due to our limited sample size. But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition. This corresponds with the timing when new crops spread into southwest China. Despite of no C4 plant origin in our samples, one potsherd implies a source of C4 plant-fed ruminant. In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife. This might be related to local ecological zones and environments where wild animals were easier to obtain. Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.

-adoption of farming in a relatively enclosed circumstances?

# Conclusion

Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds. The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China. It is most likely the terrestrial plant signal from C3 plant, implying rice processing in the pot. However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains. Instead, food processing in the early Neolithic is mostly ruminant fats that could originate from local wild animals, such as deer. This indicates that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming. We recognize the limitation of our interpretation of the local diet due to the small sample size. However, our study can still offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

# Acknowledgements

>>>>

---
title: "Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [UofO]
      correspondence: true
  - Kuei-Chen Lin:
      email: kueichenlin@asihp.net
      institute: [UofA]
      correspondence: true
institute:
  - UofO: Institute of History and Philology, Academia Sinica 
  - UofA: Institute of History and Philology, Academia Sinica 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or species across areas during the Neolithic transition is one of the important archaeological issues to understand the process of adoption of new practices. It is widely accepted that mixed farming of rice/millet was introduced into Southwest China from Central and Northern China around 5000 BP. However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear. The excavation of Guijiabao (ca. 5000-800 BP) in the Yanyuan basin in Southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarkers identification and compounds-specific stable isotope analysis. Our results show the signals of terrestrial animals for most potsherds with one showing a possible source from C3 plants during the Neolithic transition. Our case study indicates hunting might be the primary subsistence, supplemented by rice/millet farming, and also demonstrates a perspective from organic residue analysis to explore food consumption during the Neolithic period. A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diet and subsistence.

keywords: |
  Lipid residue analysis; Ceramics; Southwest China; Stable isotope analysis; Food practices
highlights: |
  These are the highlights. 
---

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022]. Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022; @Weisskopf2015]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.

@@

The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies or culinary practices across a region or over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021]. This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020]. We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture through our case study.

@@

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. The common ceramic styles in the early Neolithic period include grayish high-tempered fine ware and reddish brown coarse ware with surface decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Among the ceramics, incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a]. About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016]. This coresponds with a C4 based diet based on isotopic analysis on four human bones [@Lin2022]. The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022]. A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 BP [@Hao2022]. Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

<!--info about Gaoshan-->
To have a comparison, we also look at a sample from Gaoshan in southwestern Chengdu Plain, which is an alluvial plain located between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013; @Gaoshan2017]. A large quantity of carbonized rice grains and isotopic analysis results suggests that rice is the main crop supplemented by millets [@JiangGaoshan2017; @Lee2020]. This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows the predominance of domesticated pigs, with a small number of domesticated dogs [@He2020]. The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation.

Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP, what their proportions are in local culinary practices during the early time remains unclear. For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in the neighboring sites, such as Gaoshan, Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013]. The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.

@@

knitr::kable(Ch_sherds,
             caption = "A list of the pottery sherds analyzed in this study. The sherd labeld with 'SYG' means potsherds from Guijiabao, while the label 'CDG' refers to Gaoshan")
```

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of organic residues since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For the sample processing, we cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS. Internal standards were added before and after the total lipid extraction.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degrees C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen streams. About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). Helium carrier is in constant flow rate of 1.3 ml/min. The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 min. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode. The carrier gas was helium at 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was held at 90°C for 1.3 minute, and then raised to 320°C at a rate of 5°C/min. The final temperature was held for 22 minutes. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface. The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set from 90 °C for 1.5 minutes, and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.

@@

knitr::kable(quant_FA_tidy,
             caption = "Quantitative data of fatty acid (FA) composition (μg) for each pottery sample based on GC-FID analysis")
```

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of the ceramic or a different use. This sample also has charred interior surface, indicating its function to be associated with food processing. Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to the early Neolithic period contains a rich composition of long-chain fatty acids, reflecting good preservation. We recognize limitations and challenges for interpreting lipid source over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plants. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

@@

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijiabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). However, we are unable to determine whether there is a shift in diet over time due to our limited sample size. But the overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition. This corresponds with the timing when new crops were introduced into southwest China. Despite no C4 plant origin was found, one potsherd shows the signal of C4 plant-fed ruminant. In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife. This might be related to local ecological zones and environments where wild animals were easier to obtained compared widely adoption of agriculture. Since zooarchaeological analysis and studies of other artifacts such as stone tools are still in progress, we could only rely on preliminary excavation reports with preliminary results of other lines of evidence. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.

- discuss the conflict with the results of isotopic analysis for human bones

# Conclusion
Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds. The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China. It is most likely the terrestrial plant signal from C3 plant, implying rice processing in the pot. However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains. Instead, food processing in the early Neolithic is mostly ruminant fats that could originate from local wild animals, such as deer. This indicates that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming. We recognize the limitation of our interpretation of the local diet due to the small sample size. However, our study can still offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period. A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

# Acknowledgements
We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples. We thank Julian Sachs in the Department of Oceanography, University of Washington for his support to let us use their laboratory for sample preparation and conducting lipid extraction and analysis. We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines. We also want to express our gratitude to the anonymous reviewers for their comments that help us to improve this paper.
